The universal property of direct integrals in W$^*$-categories

### 2018-10-21 direct_integrals.tex
minor edits and corrections

% categories
\newcommand{\cat}[1]{\mathsf{#1}}		% generic category
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Vect}[1]{\mathsf{Vect}_{#1}}	% vector spaces over a field
\newcommand{\Ban}{\mathsf{Ban}}			% (complex) Banach spaces
\newcommand{\Hilb}{\mathsf{Hilb}} 		% Hilb, cat of Hilbert spaces
\newcommand{\Rep}[1]{\mathsf{Rep}(#1)}		% cat normal Hilbert space reps of a C$^*$-algebra
\newcommand{\NRep}[1]{\mathsf{nRep}(#1)}	% cat of normal Hilbert space reps of a W$^*$-algebra
\newcommand{\id}[1]{1_{#1}}			% identity morphism
\newcommand{\Kay}{\mathcal{K}} % I use a convention that the name of a letter spelt phonetically signifies its mathcal version. But semantically, this is a compact class in a measure space.
\newcommand{\powerset}{\mathcal{P}}

@@

A measure space is \emph{localizable} $\mu$ is semi-finite and $\Sigma$ modulo the null sets is complete as a Boolean algebra, which means that every collection of measurable sets has an \emph{essential supremum}. Localizable measure spaces were introduced by Segal \cite{segal} as a characterization of the measure spaces such that $L^\infty(X)$, the C$^*$-algebra of essentially bounded measurable functions modulo null sets, is a W$^*$-algebra.

@@

We can shed some light on the notion of localizability with some simple counter-examples based on the same underlying $\sigma$-algebra. Let $X$ be any uncountable set, and take $\Sigma$ to be the sets that are either countable or have countable complement (these sets are called \emph{cocountable}). This is easily seen to be a $\sigma$-algebra. We make $(X,\Sigma)$ into a measure space by three different choices of measure, each of which illustrates different features. We define the counting measure $\kappa : \Sigma \rightarrow [0,\infty]$ by
\begin{equation}
\label{CountingMeasDefnEqn}
\kappa(S) = \begin{cases} |S| & \text{if } S \text{ is finite} \\
                          \infty & \text{if } S \text{ is infinite} \end{cases}.
\end{equation}
This is a countably additive measure, and its only null set is $\emptyset$. The space $(X,\Sigma,\kappa)$ is semi-finite, but not localizable, because $\Sigma$ is not a complete Boolean algebra (otherwise it would contain all subsets of $X$).

We define the measure $\mu_c : \Sigma \rightarrow [0,1]$ by
\begin{equation}
\label{CountableCocountableMeasDefnEqn}
\mu_c(S) = \begin{cases} 0 & \text{if } S \text{ is countable} \\
                         1 & \text{if } S \text{ is cocountable} \end{cases}.
\end{equation}
Then $(X,\Sigma,\mu_c)$ is a probability space, and therefore localizable. In fact, we can see that $\Sigma/ \Null$ is a 2-element Boolean algebra, equipped with its unique probability measure.

We define the measure $\mu_\infty : \Sigma \rightarrow [0,\infty]$ by
\[
\mu_\infty(S) = \begin{cases} 0 & \text{if } S \text{ is countable} \\
                             \infty & \text{if } S \text{ is cocountable} \end{cases}.
\]

@@

Unfortunately, there exist localizable measure spaces $X$, $Y$ such that $L^\infty(X) \cong L^\infty(Y)$ but there is no measurable map from $X$ to $Y$ taking nullsets to nullsets under preimage. For example, if we consider $(1,\powerset(1),\nu)$, where $\nu$ is the unique probability measure, and the space $(X,\Sigma,\mu_c)$ from \eqref{CountableCocountableMeasDefnEqn}. A map $f: 1 \rightarrow X$ that preserves nullsets under preimage cannot exist, because the preimage of every singleton is required to be the empty set. But $1$ and $X$ both have $2$ as their measure algebra. This is enough to show that $L^\infty(1) \cong L^\infty(X)$, but it can also be seen directly by showing that for each measurable function $f : X \rightarrow \C$ is constant on some cocountable set, so $L^\infty(X) \cong \C \cong L^\infty(1)$.

To get out of this difficulty, we restrict ourselves to \emph{complete strictly localizable compact} measure spaces. Doing so still allows us to use all commutative W$^*$-algebras, which would not be the case if we restricted to $\sigma$-finite measures on standard Borel spaces, for example.

To define these notions, we first define the finite intersection property, and what a compact class is. Let $X$ be a set, and $\Kay \subseteq \powerset(X)$ be a family of sets. We say it has the \emph{finite intersection property} if for each finite subfamily $(K_i)_{i \in I}$, $K_i \in \Kay$, we have $\bigcap_{i \in I} K_i \neq 0$. We say that $\Kay$ is a \emph{compact class} if for any subfamily $\Kay' \subseteq \Kay$ having the finite intersection property, $\bigcap \Kay' \neq \emptyset$. For example, the closed sets of any compact topological space form a compact class. Slightly less trivially, the compact subsets of any Hausdorff space form a compact class.

>>>>

% categories
\newcommand{\cat}[1]{\mathsf{#1}}		% generic category
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Vect}[1]{\mathsf{Vect}_{#1}}	% vector spaces over a field
\newcommand{\Ban}{\mathsf{Ban}}			% (complex) Banach spaces
\newcommand{\Hilb}{\mathsf{Hilb}} 		% Hilb, cat of Hilbert spaces
\newcommand{\Rep}[1]{\mathsf{Rep}(#1)}		% cat normal Hilbert space reps of a C$^*$-algebra
\newcommand{\NRep}[1]{\mathsf{nRep}(#1)}	% cat of normal Hilbert space reps of a W$^*$-algebra
\newcommand{\id}[1]{1_{#1}}			% identity morphism
\newcommand{\Kay}{\mathcal{K}} % I use a convention that the name of a letter spelt phonetically signifies its mathcal version. But semantically, this is a compact class in a measure space.
\newcommand{\powerset}[1]{\mathcal{P}(#1)}

@@

A measure space is \emph{localizable} if $\mu$ is semi-finite and $\Sigma$ modulo the null sets is complete as a Boolean algebra, which means that every collection of measurable sets has an \emph{essential supremum}. Localizable measure spaces were introduced by Segal \cite{segal} as a characterization of the measure spaces such that $L^\infty(X)$, the C$^*$-algebra of essentially bounded measurable functions modulo null sets, is a W$^*$-algebra.

@@

We can shed some light on the notion of localizability with some simple counter-examples based on the same underlying $\sigma$-algebra. Let $X$ be any uncountable set, and take $\Sigma$ to be the sets that are either countable or have countable complement (these sets are called \emph{cocountable}). This is easily seen to be a $\sigma$-algebra. We make $(X,\Sigma)$ into a measure space by three different choices of measure, each of which illustrates different features. We define the counting measure $\kappa : \Sigma \rightarrow [0,\infty]$ by
\begin{equation}
\label{CountingMeasDefnEqn}
\kappa(S) = \begin{cases} |S| & \text{if } S \text{ is finite}, \\
                          \infty & \text{if } S \text{ is infinite}. \end{cases}
\end{equation}
This is a countably additive measure, and its only null set is $\emptyset$. The space $(X,\Sigma,\kappa)$ is semi-finite, but not localizable, because $\Sigma$ is not a complete Boolean algebra (otherwise it would have to contain all subsets of $X$).

We define the measure $\mu_c : \Sigma \rightarrow [0,1]$ by
\begin{equation}
\label{CountableCocountableMeasDefnEqn}
\mu_c(S) = \begin{cases} 0 & \text{if } S \text{ is countable}, \\
                         1 & \text{if } S \text{ is cocountable}. \end{cases}
\end{equation}
Then $(X,\Sigma,\mu_c)$ is a probability space, and therefore localizable. In fact, we can see that $\Sigma/ \Null$ is a 2-element Boolean algebra, equipped with its unique probability measure.

We define the measure $\mu_\infty : \Sigma \rightarrow [0,\infty]$ by
\[
\mu_\infty(S) = \begin{cases} 0 & \text{if } S \text{ is countable}, \\
                             \infty & \text{if } S \text{ is cocountable}. \end{cases}
\]

@@

Unfortunately, there exist localizable measure spaces $X$, $Y$ such that $L^\infty(X) \cong L^\infty(Y)$ but there is no measurable map from $X$ to $Y$ taking nullsets to nullsets under preimage. For example, if we consider $(1,\powerset{1},\nu)$, where $\nu$ is the unique probability measure, and the space $(X,\Sigma,\mu_c)$ from \eqref{CountableCocountableMeasDefnEqn}. A map $f: 1 \rightarrow X$ that preserves nullsets under preimage cannot exist, because the preimage of every singleton is required to be the empty set. But $1$ and $X$ both have $2$ as their measure algebra. This is enough to show that $L^\infty(1) \cong L^\infty(X)$, but it can also be seen directly by showing that for each measurable function $f : X \rightarrow \C$ is constant on some cocountable set, so $L^\infty(X) \cong \C \cong L^\infty(1)$.

In order to avoid pitfalls like this, we restrict ourselves to \emph{complete strictly localizable compact} measure spaces. Doing so still allows us to use all commutative W$^*$-algebras, which would not be the case if we restricted to $\sigma$-finite measures on standard Borel spaces, for example.

To define these notions, we first define the finite intersection property, and what a compact class is. Let $X$ be a set, and $\Kay \subseteq \powerset{X}$ be a family of sets. We say it has the \emph{finite intersection property} if for each finite subfamily $(K_i)_{i \in I}$, $K_i \in \Kay$, we have $\bigcap_{i \in I} K_i \neq 0$. We say that $\Kay$ is a \emph{compact class} if for any subfamily $\Kay' \subseteq \Kay$ having the finite intersection property, $\bigcap \Kay' \neq \emptyset$. For example, the closed sets of any compact topological space form a compact class. Slightly less trivially, the compact subsets of any Hausdorff space form a compact class.
